---
title: "lab_10"
author: "Alexis Saldivar (PID: A69038083)"
format: pdf
---

## Load Data
```{r}
library(skimr)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(corrplot)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy.data <- read.csv(url, row.names = 1)
```

## Summary
```{r}
n_type <- ncol(candy.data)
fruity <- sum(candy.data["fruity"])
```
> Q1. How many candy types are in the dataset?

There are ``r n_type` candy types.

> Q2. How many fruity candy types are in the datset?

`r fruity`

```{r}
my_winp <- candy.data["Hershey's Kisses", "winpercent"]
kitkat_winp <- candy.data["Kit Kat", "winpercent"]
tootsie_winp <- candy.data["Tootsie Roll Snack Bars", "winpercent"]
```

> Q3.  What is your favorite candy in the dataset and what is it’s winpercent value?

Hershey's Kisses: `r my_winp`

> Q4. What is the winpercent value for “Kit Kat”?

`r kitkat_winp`

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

`r tootsie_winp`

```{r}
skim(candy.data)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the winpercent column

> Q7. What do you think a zero and one represent for the candy$chocolate column?

If a given candy is (1) or isn't (0) of type chocolate

> Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy.data) +
  aes(winpercent) +
  geom_histogram()
```
> Q9. Is the distribution of winpercent values symmetrical?

No, it's biased to the left

> Q10. Is the center of the distribution above or below 50%?

Below 50%

```{r}
chocolate <- candy.data |> filter(chocolate==1) |> select(winpercent)
ch_avg <- mean(chocolate$winpercent)

fruity <- candy.data |> filter(fruity==1) |> select(winpercent)
fr_avg <- mean(fruity$winpercent)

ans <- t.test(chocolate$winpercent, fruity$winpercent)
ans$p.value

candy.data |> arrange(-winpercent) |> head(5)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy is higher ranked `r ch_avg` than fruity `r fr_avg`

> Q12. Is this difference statistically significant?

Yes the difference is statistically significant p-value = `r ans$p.value`

> Q13. What are the five least liked candy types in this set?

`r candy.data |> arrange(winpercent) |> head(5)`

> Q14. What are the top 5 all time favorite candy types out of this set?

`r candy.data |> arrange(-winpercent) |> head(5)`

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy.data) +
  aes(winpercent, rownames(candy.data)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy.data) +
  aes(winpercent, reorder(rownames(candy.data), winpercent)) +
  geom_col()
```
```{r}
my_cols=rep("black", nrow(candy.data))
my_cols[as.logical(candy.data$chocolate)] = "chocolate"
my_cols[as.logical(candy.data$bar)] = "brown"
my_cols[as.logical(candy.data$fruity)] = "pink"

ggplot(candy.data) + 
  aes(winpercent, reorder(rownames(candy.data),winpercent)) +
  geom_col(fill=my_cols) 
```
> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

```{r}
ggplot(candy.data) +
  aes(winpercent, pricepercent, label=rownames(candy.data)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Chocolate - Reese's Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The least popular is Nik L Nip

```{r}
candy.data |> arrange(-pricepercent) |> head(5) |> select(winpercent) |> arrange(winpercent)
```

```{r}
cij <- cor(candy.data)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

```{r}
candy.pca <- prcomp(candy.data, scale=T)
summary(candy.pca)
```
```{r}
ggplot(candy.pca$x) +
  aes(PC1, PC2) +
  geom_point(col=my_cols) +
  geom_text_repel(label=rownames(candy.pca$x))
```

```{r}
par(mar=c(8,4,2,2))
barplot(candy.pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, pluribus.













